Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-05-26 bwamem.tex
The version to appear in arXiv as the 2nd version
The supplementary is for a record only. Some of the content has been moved back
to the preprint.

\begin{figure}[tb]
\centering
\includegraphics[width=0.4\textwidth]{alnroc-se}\\
\includegraphics[width=0.4\textwidth]{alnroc-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. Alignments with mapping quality 3 or lower
are excluded. An alignment is \emph{wrong} if after correcting clipping, its
start position is at least 5bp away from the simulated position. $10^6$ pairs
of 101bp reads are simulated from the human reference genome using wgsim
(http://bit.ly/wgsim2) with 1.5\% substitution errors and 0.2\% indel variants.
The insert size follows a normal distribution $N(500,50^2)$. The reads are
aligned back to the genome either as single end (SE; top panel) or as paired
end (PE; bottom panel). GEM is configured to allow up to 5 gaps and to output
suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb'
for PE). GEM does not compute mapping quality. Its mapping quality is estimated
with a BWA-like algorithm with suboptimal
alignments available.  Other mappers are run with the default setting except
for specifying the insert size distribution. The run time in seconds on a
single CPU core is shown in the parentheses.}\label{fig:eval}
\end{figure}

@@

We evaluated the performance of BWA-MEM on simulated data together with
NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, BWA-SW and BWA
(Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM is close
to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.  On
speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
6 times as fast as Bowtie2 and Cushaw2 for a 650bp long-read data set.

@@

%To evaluate the performance for even longer query sequences, we aligned the
%{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
%the 536 strain (AC:NC\_008253) with both BWA-MEM and
%nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
%giving 105,505 substitution differences between the strains. BWA-MEM is slower.
%It finished the alignment in 131 seconds, reporting 104,321 substitutions of
%which 102,241 overlapping with the nucmer output. Manual examination of the
%substitutions unique to each aligner reveals that most of them fall in short
%regions with very high divergence. It is unclear which aligner is better. Note
%that although nucmer is faster in the evaluation, only BWA-MEM scales well to
%large genomes.

>>>>

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{alnroc-color-se}\\
\includegraphics[width=0.5\textwidth]{alnroc-color-pe}
\caption{Percent mapped reads as a function of the false alignment rate under
different mapping quality cutoff. Alignments with mapping quality 3 or lower
are excluded. An alignment is \emph{wrong} if after correcting clipping, its
start position is within 20bp from the simulated position. $10^6$ pairs
of 101bp reads are simulated from the human reference genome using wgsim
(http://bit.ly/wgsim2) with 1.5\% substitution errors and 0.2\% indel variants.
The insert size follows a normal distribution $N(500,50^2)$. The reads are
aligned back to the genome either as single end (SE; top panel) or as paired
end (PE; bottom panel). GEM is configured to allow up to 5 gaps and to output
suboptimal alignments (option `--e5 --m5 --s1' for SE and `--e5 --m5 --s1 --pb'
for PE). GEM does not compute mapping quality. Its mapping quality is estimated
with a BWA-like algorithm with suboptimal alignments available. Other mappers
are run with the default setting except for specifying the insert size
distribution.
%GSNAP and STAR essentially do not report mapping quality, so they are not shown in the plot.
%For SE, GSNAP's accuracy is placed at $(2.8\times10^{-3},96.1\%)$ in the plot and STAR at $(4.6\times10^{-3},95.5\%)$.
%For PE, GSNAP achieves $(1.1\times10^{-3},97.8\%)$ accuracy and STAR $(2.0\times10^{-3},96.8\%)$.
The run time in seconds on a single CPU core is shown in the
parentheses.}\label{fig:eval}
\end{figure}

@@

We evaluated the performance of BWA-MEM on simulated data together with
NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, SeqAlto~\citep{Mu:2012fk}, %GSNAP~\citep{Wu:2010uq}, STAR~\citep{Dobin:2013kx},
BWA-SW
and BWA (Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM
is close to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.
On speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
6 times as fast as Bowtie2 and Cushaw2 for a 650bp long-read data set.

@@

To evaluate the performance for even longer query sequences, we aligned the
{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
the 536 strain (AC:NC\_008253) with both BWA-MEM and
nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
giving 105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of
which 102,241 overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short
regions with very high divergence. It is unclear which aligner is better. Note
that although nucmer is faster in the evaluation, only BWA-MEM scales well to
large genomes.
